---
title: "NEONATAL AND UNDER5 MORTALITY RATE"
author: "Iyvonne Anne Ayoyi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Data Analysis Projects/CEMA DATA SCIENCE")
```

Loading Libraries
```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(sf)
library(viridis)
```

Load and Prepare Data
```{r}
# Load WHO child mortality dataset (assumes combined data file)
data <- read_csv("dataset_datascience.csv")

# View column names to verify structure
colnames(data)
```

```{r}
# Load shapefile for mapping, standardize names
eac_shape <- st_read("EAC_COUNTRIES.shp") %>%
  mutate(Country = case_when(
    NAME == "Democratic Republic of the Congo" ~ "Congo DRC",
    NAME == "United Republic of Tanzania" ~ "Tanzania",
    TRUE ~ NAME
  ))
print(unique(eac_shape$Country))
```



```{r}
# Define EAC countries
eac_countries <- c("Burundi", "Kenya", "Rwanda", "South Sudan", 
                   "Tanzania", "Uganda", "Congo DRC", "Somalia")

# Standardize and filter for EAC countries only
mortality_data <- data %>%
  mutate(
    Country = case_when(
      `Geographic area` %in% c("Congo, Dem. Rep.", "Democratic Republic of the Congo") ~ "Congo DRC",
      `Geographic area` %in% c("Tanzania, United Rep.", "United Republic of Tanzania") ~ "Tanzania",
      TRUE ~ `Geographic area`
    ),
    Year = as.numeric(substr(`Series Year`, 1, 4))
  ) %>%
  filter(Country %in% eac_countries)
mortality_data

```

```{r}
unique(data$Indicator)
```


```{r}
# Separate neonatal and under-5 mortality
neonatal <- mortality_data %>%
  filter(Indicator == "Neonatal mortality rate")

under5 <- mortality_data %>%
  filter(Indicator == "Under-five mortality rate")


summary(neonatal)
summary(under5)

```




Merge with Latest Data and Add Labels
```{r}
#Latest data
get_latest <- function(df) {
  df %>%
    group_by(Country) %>%
    filter(Year == max(Year, na.rm = TRUE)) %>%
    ungroup() %>%
    select(Country, `Observation Value`, Year)
}

neonatal_latest <- get_latest(neonatal) %>% rename(NeonatalRate = `Observation Value`)
under5_latest <- get_latest(under5) %>% rename(Under5Rate = `Observation Value`)
neonatal_latest
under5_latest
```


```{r}
# Merge with shapefile
merge_shape <- function(shape, data, rate_col) {
  df <- left_join(shape, data, by = "Country")
  centroid_coords <- st_centroid(df) %>% st_coordinates() %>% as.data.frame()
  df$lon <- centroid_coords$X
  df$lat <- centroid_coords$Y
  df$Label <- ifelse(is.na(df[[rate_col]]), paste0(df$Country, "\nNA"), df$Country)
  return(df)
}

neonatal_map <- merge_shape(eac_shape, neonatal_latest, "NeonatalRate")
under5_map <- merge_shape(eac_shape, under5_latest, "Under5Rate")
```


#Plotting Choropleth Maps

```{r}
plot_choropleth <- function(df, rate_col, title) {
  ggplot(df) +
    geom_sf(aes(fill = .data[[rate_col]]), color = "white") +
    geom_text(aes(x = lon, y = lat, label = Label), size = 3, color = "black") +
    scale_fill_viridis_c(
      name = "Deaths/1,000",
      option = "C",
      na.value = "lightgrey",
      direction = -1,  
  breaks = pretty(range(df[[rate_col]], na.rm = TRUE), n = 5)
    ) +
    labs(title = title) +
    theme_minimal()+
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "right"
    )
}

plot_choropleth(neonatal_map, "NeonatalRate", "Latest Neonatal Mortality in EAC")
plot_choropleth(under5_map, "Under5Rate", "Latest Under-Five Mortality in EAC")
```
```



### Analyzing Trend Over Time

```{r}
#Preparing Trend Data
prepare_trend <- function(df, rate_col) {
  df %>%
    group_by(Year) %>%
    mutate(AvgRate = mean(.data[[rate_col]], na.rm = TRUE)) %>%
    ungroup()
}
# Call the function with the correct rate column for each dataset
neonatal_trend <- prepare_trend(neonatal, "Observation Value")
under5_trend   <- prepare_trend(under5, "Observation Value")

neonatal_trend
under5_trend

```


```{r}
plot_trends <- function(df, rate_col, title) {
  ggplot(df, aes(x = Year)) +
    geom_line(aes(y = AvgRate), color = "#0072B2", linetype = "dashed", size = 1.2) +
    geom_jitter(aes(y = .data[[rate_col]], color = Country), width = 0.3, size = 2, alpha = 0.6) +
    scale_color_viridis_d(name = "Country") +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    labs(title = title, y = "Deaths/1,000 live births", x = "Year") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
}
plot_trends(neonatal_trend, "Observation Value", "Neonatal Mortality Trends in EAC")
plot_trends(under5_trend, "Observation Value", "Under-Five Mortality Trends in EAC")

```



Finding Country with the Highest Mortality Rates in EAC
```{r}
#Highest Under 5
cat("Country with Highest Under-5 Mortality:\n")
under5_latest %>% 
  arrange(desc(Under5Rate)) %>% 
  select(Country, Under5Rate,Year)
```


```{r}
#Highest Neonatal
cat("\nCountry with Highest Neonatal Mortality:\n")
neonatal_latest %>% 
  arrange(desc(NeonatalRate)) %>% 
  select(Country, NeonatalRate, Year)
```
